Fail Faster

### 2025-03-05 paper/paper.tex
move structure around a bit more
progress

@@

@@

let bind (g : 'a gen) (k : 'a -> 'b gen) : 'b gen =
  fun size_c random_c ->
    .<
      let x = .~(g size_c random_c) in
      .~(
        (k .<x>.) size_c random_c
      )
    >.

@@

\begin{lstlisting}
let int_pair_staged : ((int * int) code) gen =
  let%bind x = int 0 100 in
  let%bind y = int 0 x in
  return (x,y)

let int_pair : (int * int) Bq.t code = to_bq int_pair_staged
====
.<
  Bq.create (fun size random ->
    let x = SR.int random 0 100 in
    let y = SR.int random 0 x in
    (x,y)
  )
.>
\end{lstlisting}

@@

\begin{lstlisting}
let grades : charBq.t gen = to_bq (
  weighted_union [
    (.<3.0>., 'a');
    (.<2.0>., 'b');
    (.<1.0>., 'c');
  ]
)
===
Bq.create (
  fun size random ->
    let sum = 3.0 + 2.0 + 1.0 in
    let r = SR.int random 0 sum in
    if r <= 3.0 then 'a'
    else
      let r' = r - 3.0 in
      if r' <= 2.0 then 'b'
      else
        let r'' = r' - 2.0 in
        if r'' <= 1.0 then 'c'
        else
          failwith "Error"
)
\end{lstlisting}

\subsection{``The Trick'' and Effect Ordering}

\jwc{Note: tradeoff between maintaining the effect ordering and even more performance. }

\jwc{Note with bind that we don't split because it's imperatively updated. Note that this omits the need to use SR, but we use it to maintain backwards compat.}

\subsection{Comparing Generators}

\jwc{NOTE: we should test generator speed across both languages, but speed -> bugfinding ability in only OCaml.}
\jwc{
  Baseline generators to test speed in both languages:
  \begin{itemize}
    \item Single int
    \item Pair of ints, constrained
    \item List of ints
  \end{itemize}
}

\subsubsection{Existing Approaches}
\subsubsection{Equality up to Program Equality}
\subsubsection{Effect Ordering}
\subsubsection{PBT for PBT}
\subsubsection{Does maintaining program equality in \name\ affect performance?}
\section{Evaluation}
\subsubsection{Implementation}
\subsection{Benchmarking speed \& resource usage}
\subsection{Impact on bug-finding ability}
\section{Conclusion \& Future Work}
%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{bib}

>>>>

\jwc{ofc, sometimes you only know the list at runtime (see the STLC generator for an example), so you the library also includes this.}

@@

\jwc{Note that some of this is a lie, but it's a good first cut: point to the section on \emph{The Trick} to see the real version.}

@@

let bind (g : 'a gen) (k : 'a code -> 'b gen) : 'b gen =
  fun size_c random_c ->
    .<
      let x = .~(g size_c random_c) in
      .~(
        (k .<x>.) size_c random_c
      )
    >.

@@

\begin{lstlisting}
let int_pair_staged : ((int * int) code) gen =
  bind (int .<0>. .<100>.) (fun cx ->
    bind (int .<0> cx) (fun cy ->
      return .<(.~cx,.~cy)>.
    )
  )
let int_pair : (int * int) Bq.t code = to_bq int_pair_staged
====
.<
  Bq.create (fun size random ->
    let x = SR.int random 0 100 in
    let y = SR.int random 0 x in
    (x,y)
  )
.>
\end{lstlisting}

@@

\begin{lstlisting}
let grades : charBq.t gen = to_bq (
  weighted_union [
    (.<3.0>., .<'a'>.);
    (.<2.0>., .<'b'>.);
    (.<1.0>., .<'c'>.);
  ]
)
===
Bq.create (
  fun size random ->
    let sum = 3.0 + 2.0 + 1.0 in
    let r = SR.int random 0 sum in
    if r <= 3.0 then 'a'
    else
      let r' = r - 3.0 in
      if r' <= 2.0 then 'b'
      else
        let r'' = r' - 2.0 in
        if r'' <= 1.0 then 'c'
        else
          failwith "Error"
)
\end{lstlisting}

\subsection{Effect Ordering, and Equivalence is Program Equality}
\jwc{``Wait, why was the defintion of bind like that''?
Careful readers might note that the definition of bind was a little strange. In particular, why not just do this?}
\begin{lstlisting}
  let bind (g : 'a gen) (k : 'a code -> 'b gen) : 'b gen =
    fun size_c random_c ->
      k (g size_c random_c) size_c random_c
\end{lstlisting}
\jwc{The answer is that this doesn't generate the right code! In particular, this definition of bind means that
\texttt{bind (int 0 1) (fun x -> return .<(.~x,.~x)>.)} would generate \texttt{fun size random -> (int random 0 1, int random 0 1)},
which is \emph{not} the desired behavior. (Cite cbn/cbv randomness here).
A splice is a (hygenic) syntactic replacment, so
if a function \texttt{f : 'a code -> 'b code} splices its argument multiple times into its output, the argument code will be directly copied in.
This is why we insert a let-binding in the definition of bind, to ensure that
the explicit sequencing of bind is preserved into explicit sequencing in the
generated code.
}

\jwc{
The library is carefully design to preserve exactly the effect order of BQ. If
you write the same generator in both, they run identically: for the same size
and random seed, they produce the same value.
}

\jwc{Note: tradeoff between maintaining the effect ordering and even more performance.}

\subsection{A Real Monad Instance, CPS, and ``The Trick''}

\jwc{
  \begin{itemize}
    \item Careful readers might note that the types of return and bind are not actually the right types.
    \item This means that they aren't compatible with ppx-jane syntax.
    \item We could try a different type, \texttt{type 'a gen = int code -> SR.t code -> 'a}. This has an actual monad instance (show), and ou can write teh combinators of before, with \texttt{'a code gen} everywhere you had \texttt{'a gen} before.
    \item But because the result type isn't always in \texttt{code}, \texttt{bind} can't perform the let-insertion needed to preserve effect ordering.
    \item Instead, we turn to a classic technique from the multistage programming literature: writing our staged program in continuation-passing style.
    \item We define \texttt{type 'a codecps = 'z. ('a -> 'z code) -> 'z code}. This is the polymorphic CPS monad, with the continuation type always in `code'.
    (Cite: ``improving binding times without explicit cps conversion'', ``Multi-stage programming with functors and monads: Eliminating abstraction overhead from generic code'', and ``Closure-Free Functional Programming in a Two-Level Type Theory'')
    \item This lets us define the function \texttt{let\_insert : 'a code -> 'a codecps} as \texttt{let\_insert cx = fun k -> k .< let x = .~cx in .~(k .<x>.) >.}.
    \item We then define \texttt{type 'a gen = int code -> SR.t code -> 'a codecps}, which lets us have it both ways. the effectful programs like \texttt{int} do their \emph{own} let-insertion
  \end{itemize}
}

\jwc{This gives us the final definitions:}

\begin{lstlisting}
module CodeCps = struct
  type 'a t = 'z. ('a -> 'z code) -> 'z code

  let run (t : 'a code t) : 'a code = t (fun x -> x)

  let let_insert (cx : 'a code) : 'a code t =
    fun k -> k .< let x = .~cx in .~(k .<x>.) >.
end

module Gen = struct
  type 'a t = int code -> SR.t code -> 'a CodeCps.t

  let return (x : 'a) : 'a t = fun _ _ -> x

  let bind (g : 'a t) (f : 'a -> 'b t) =
    fun size random ->
      f (g size random) size random

  let int (lo : int code) (hi : int code) : int code t =
    fun size random ->
      let_insert .< SR.int .~random .~lo .~hi >.
end
\end{lstlisting}

\jwc {
And this has the desired effect that
\texttt{bind (int .<0>. .<1.>) (fun x -> return .<(.~x,.~x)>.)} generates
\texttt{fun size random -> let x = SR.int random 0 1 in (x,x)}
}

\subsubsection{The Trick}

\subsection{Staged Type-Derived Generators}

\jwc{Todo: Thia}

\section{Evaluation}

\subsubsection{Implementation}
\subsection{Benchmarking speed \& resource usage}

\jwc{NOTE: we should test generator speed across both languages, but speed -> bugfinding ability in only OCaml.}
\jwc{
  Baseline generators to test speed in both languages:
  \begin{itemize}
    \item Single int
    \item Pair of ints, constrained
    \item List of ints
  \end{itemize}
}
\subsection{Impact on bug-finding ability}
\section{Conclusion \& Future Work}
%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{bib}
